---
title: "model 1b smoke alarm presence"
author: "Xianghui Dong"
date: "2016年1月31日"
output: html_document
mathjax: null
---

## To Get the target variable 1b: % of fire incidents within a region that didn’t have smoke detectors

[github issue](https://github.com/home-fire-risk/smoke_alarm_models/issues/12)
[model document](https://docs.google.com/document/d/1oJN-QwLVqFHOvrRNtW2KEAkNZ-PuFiqTwa8y3iXx1Sg/edit)

```
Measure 2: model  (Jonathan / Andrew)
target variable: % of fire incidents within a region that didn’t have smoke detectors (Xianghui create variable, push to J / A for model)
all fire incidents in a region(tract), how many didn't have smoke detectors? 
```

By region we are refering census tract now. 
For each tract, count
- all structured fire incidents
- the ratio of smoke detector presence in these incidents

Refer to [My previous analysis of smoke alarm presence information in NFIRS data](AlarmPresenceNFIRS.html) for background. 
1. Using structured fire as the most relevant approximation to home fires.
2. The basic module `DET_ALERT` is focused on whether the smoke detectors alerted occupants, which have lots of unknowns. Structured fire module `DETECTOR` just ask whether the smoke detectors is present, in many case it's known even whether it alerted occupants is unknown.
3. We need to further focus on smoke alarms in all possible fire detectors.
4. In general, a big portion of all fire incidents don't have smoke alarm presence information. It need much caution before using it as a indicator.

### Read data

NFIRS original cdrom data in 'e:/Data/NFIRS/'
Read original data, save as R data object.
```{r Read NFIRS data}
# ---- basic incident ----
#data_year = 2013
save_csv_basic = function(data_year){
  basic_file = paste0('e:/Data/NFIRS/', data_year, '/', data_year,'basicincident.txt')
  basic_incident = fread(basic_file, sep = '^', header = TRUE,
                  stringsAsFactors=FALSE, verbose = TRUE,
                  colClasses=list(character=c("INC_DATE", "ALARM", "ARRIVAL", "INC_CONT", "LU_CLEAR")))
  setkey(basic_incident, STATE, FDID, INC_DATE, INC_NO, EXP_NO)
  saveRDS(basic_incident, paste0('e:/data/NFIRS/', data_year, 'basicincident.rds'))
}
save_csv_basic(2012)
save_csv_basic(2013)
#data_year = 2011
save_dbf_basic = function(data_year){
  basic_file = paste0('e:/Data/NFIRS/', data_year, '/', data_year,'basicincident.dbf')
  basic_incident = data.table(read.dbf(basic_file, as.is = TRUE))
  setkey(basic_incident, STATE, FDID, INC_DATE, INC_NO, EXP_NO)
  saveRDS(basic_incident, paste0('e:/data/NFIRS/', data_year, 'basicincident.rds'))
}
save_dbf_basic(2009)
save_dbf_basic(2010)
save_dbf_basic(2011)
# ---- fire incident ----
# data_year = 2013
save_csv_fire = function(data_year){
  fire_file = paste0('e:/Data/NFIRS/', data_year, '/', data_year,'fireincident.txt')
  fire_incident = fread(fire_file, sep = '^', header = TRUE,
                    stringsAsFactors=FALSE, verbose = TRUE, colClasses=list(character=c("INC_DATE")))
  setkey(fire_incident, STATE, FDID, INC_DATE, INC_NO, EXP_NO)
  saveRDS(fire_incident, paste0('e:/data/NFIRS/', data_year, 'fireincident.rds'))
}
save_csv_fire(2012)
save_csv_fire(2013)
# data_year = 2011
save_dbf_fire = function(data_year){
  fire_file = paste0('e:/Data/NFIRS/', data_year, '/', data_year,'fireincident.dbf')
  fire_incident = data.table(read.dbf(fire_file, as.is = TRUE))
  setkey(fire_incident, STATE, FDID, INC_DATE, INC_NO, EXP_NO)
  saveRDS(fire_incident, paste0('e:/data/NFIRS/', data_year, 'fireincident.rds'))
}
save_dbf_fire(2009)
save_dbf_fire(2010)
save_dbf_fire(2011)
```

### Calculate target variables

Geocoded incident address and the original addresses were loaded from R data object saved before.

```{r variable of percentage of fire that no smoke detectors}
# read original data, geocoding result. 
data_year = '2010'
load(paste0('data/address/', data_year, '_formated_addresses.Rdata')) # variable name: address, 21 columns
load(paste0('data/address/', data_year, '_geocoded_addresses.Rdata')) # variable name: geocoded_address, 13 columns for 2012
setkey(address, row_seq) # csv sorted, Rdata not sorted.
table(address$address_type)
# load basic and fire incident file of data_year, get incident type, link to row_seq, get location
basic_incident = readRDS(paste0('e:/data/NFIRS/', data_year, 'basicincident.rds'))
fire_incident = readRDS(paste0('e:/data/NFIRS/', data_year, 'fireincident.rds'))
# 2013: basic - 2,003,907 fire - 554,671 building_fire - 259,732
building_fire = fire_incident[basic_incident[grep('111', INC_TYPE)]]
# N for no detectors, note this is different from no smoke alarms.
table(building_fire$DETECTOR, useNA = "always")
# further break down on detector type. 
# 2, 4, 0 have detector but not smoke alarms. this and code N together count as "no smoke alarm"
# 1, 5 should indicate have smoke alarm present
table(building_fire$DET_TYPE, useNA = "always")
building_fire[, smoke_alarm_presence := NA]
building_fire[DETECTOR == 'N', smoke_alarm_presence := FALSE]
building_fire[DET_TYPE %in% c("2", "4", "0"), smoke_alarm_presence := FALSE]
building_fire[DET_TYPE %in% c("1", "5"), smoke_alarm_presence := TRUE]
table(building_fire$smoke_alarm_presence,useNA = "always")
# too many columns in building_fire. if we pick columns later, the data year difference in building_fire could cause problems for numbered column referencing. Better pick columns now, and merge geocode result later, no need to pick later.
building_fire_smoke = building_fire[, .(STATE, FDID, INC_DATE, INC_NO, EXP_NO, smoke_alarm_presence)]
# for each tract, get T/F/NA ratio, total count
# merge with geocoded addresses
setkey(address, row_seq)
setkey(geocoded_address, row_seq)
address_merged = merge(address, geocoded_address) 
# checking rating value and accuracy of geocoding
# View(address_merged[rating > 60 & rating < 70][sample(.N, 100)])
# View(address_merged[rating > 40 & rating < 50][sample(.N, 100)])
# View(address_merged[rating > 30 & rating < 40][sample(.N, 100)])
# View(address_merged[rating > 20 & rating < 30][sample(.N, 100)])
# View(address_merged[rating > 10 & rating < 20][sample(.N, 100)])
address_merged[rating >= 0 & rating < 30, .N] / address_merged[, .N]
address_merged[rating >= 0 & rating < 20, .N] / address_merged[, .N]
# some address have empty output for tabblock matching, census website with current data have internal error for them too.
good_address_merged = address_merged[rating >= 0 & rating < 21 & tractid != '', ]
# rating < 20, 74% in total are very good results. 
setkey(good_address_merged, STATE, FDID, INC_DATE, INC_NO, EXP_NO)
building_fire_smoke_geocoded = merge(building_fire_smoke, good_address_merged)
# col number depend on data structure, not a good method.
# smoke_alarm_presence is what we want, came from the newly added last column of building_fire, i.e. 117th column
# count all fire, alarm presented, not presented, Not Known for each tract
building_fire_smoke_geocoded[, c('all_fire', 'alarm_presented') 
                                      := list(.N, sum(smoke_alarm_presence, na.rm = TRUE)), 
                                      by = tractid]
all_fire = building_fire_smoke_geocoded[, .N, by = tractid]
setnames(all_fire, 'N', 'all_fire')
alarm_presented = building_fire_smoke_geocoded[, 
                                               sum(smoke_alarm_presence, na.rm = TRUE),
                                               by = tractid]
setnames(alarm_presented, 'V1', 'alarm_presented')
alarm_not_presented = building_fire_smoke_geocoded[, 
                                                        length(smoke_alarm_presence[which(smoke_alarm_presence == FALSE)]), 
                                                        by = tractid]
setnames(alarm_not_presented, 'V1', 'alarm_not_presented')
alarm_unknown = building_fire_smoke_geocoded[, 
                                                      length(smoke_alarm_presence[which(is.na(smoke_alarm_presence))]),
                                                      by = tractid]
setnames(alarm_unknown, 'V1', 'alarm_unknown')
setkey(all_fire, tractid)
setkey(alarm_presented, tractid)
setkey(alarm_not_presented, tractid)
setkey(alarm_unknown, tractid)
alarm_presence_by_tract = merge(all_fire, alarm_unknown)
alarm_presence_by_tract = merge(alarm_presence_by_tract, alarm_presented)
alarm_presence_by_tract = merge(alarm_presence_by_tract, alarm_not_presented)
saveRDS(alarm_presence_by_tract, file = paste0('results/', data_year, '_alarm_presence_by_tract.rds'))
```

### combine multiple years statistics
```{r combine multiple years statistics}
alarm_presence_years = list()
for(data_year in 2009:2013){
  current = readRDS(paste0('results/', data_year, '_alarm_presence_by_tract.rds'))
  current[, ratio_no_alarm_in_all := alarm_not_presented / all_fire]
  current[, ratio_no_alarm_in_all_known := 
            alarm_not_presented / (alarm_presented + alarm_not_presented)]
  alarm_presence_years[[toString(data_year)]] = current
  write.csv(alarm_presence_years[[toString(data_year)]], 
            paste0('results/', data_year, '_alarm_presence_by_tract.csv'),
            quote = TRUE)
}
alarm_presence_2009_2013 = rbindlist(alarm_presence_years)
alarm_presence_2009_2013[, c('all_fire_all_years', 'alarm_unknown_all_years',
                             'alarm_presented_all_years',
                             'alarm_not_presented_all_years') := 
                           list(sum(all_fire), sum(alarm_unknown),
                                sum(alarm_presented), sum(alarm_not_presented)), 
                         by = tractid]
alarm_presence_all = alarm_presence_2009_2013[,c(1, 8:11), with = FALSE]
setkey(alarm_presence_all, tractid)
alarm_presence_all = unique(alarm_presence_all)
alarm_presence_all[, ratio_no_alarm_in_all := 
                     alarm_not_presented_all_years / all_fire_all_years]
alarm_presence_all[, ratio_no_alarm_in_all_known := 
          alarm_not_presented_all_years / 
            (alarm_presented_all_years + alarm_not_presented_all_years)]
write.csv(alarm_presence_all, 'results/2009_2013_alarm_presence_by_tract.csv',
            quote = TRUE)
```
